Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-16 draft.tex
revised version

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically train a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with interactive \emph{decision tree quality plot}. %scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and described.
}

@@

(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.

@@

For example, continuous data, such as body height, is often discretized into quantile bins for visualizing age-dependent disease frequencies.

@@

	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes) and the \emph{curvature angle} ($\alpha$). %is calculated using the top, bottom and middle point.

@@

	We extracted the variables for each projection axis to assess their information gain.

@@

\noindent \textbf{Visualization of Classification Results.}

@@

Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:

@@

Furthermore, we map the number of subjects associated with a variable to the diameter in the scatter plot.

@@

\noindent \textbf{Decision Tree Visualization Quality Plot Interaction.}

@@

The analysis was performed for subjects groups of \emph{gender}, \emph{BMI} and \emph{body size} to assess their influence on the lumbar spine shape.

@@

Our method allowed to assess their discriminative power, which is largely limited to separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.

@@

Huge result spaces could be navigated fast using our interactive visualization.

@@

Dented vertebrae indicate pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.

@@

Combining the power of statistical analysis, visual analytics and classification techniques is essential for analyzing increasingly complex heterogenous population data.

@@

Our method provides a novel way to gain insight into these complex data sets and amplifies hypotheses generation.

@@

\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.

>>>>

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically train a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with a \emph{decision tree quality plot}. %scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and described.
}

@@

(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated condition.

@@

For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.

@@

	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes) and the \emph{curvature angle} ($\alpha$) for each projection axis to assess their information gain. %is calculated using the top, bottom and middle point.

@@

%	We extract the variables for each projection axis to assess their information gain.

@@

\noindent \textbf{Decision Tree Quality Plot.}

@@

Based on this we use a summary error based on the weighted mean, which incorporates the discriminative power of each manifestation and is denoted as follows:

@@

Furthermore, we map the number of subjects associated with a variable to the diameter in the decision tree quality plot.

@@

\noindent \textbf{Decision Tree Quality Plot Interaction.}

@@

The analysis was performed for subject groups of \emph{gender}, \emph{BMI} and \emph{body size} to assess their influence on the lumbar spine shape.

@@

Our method allows to assess their discriminative power, which is largely limited to separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.

@@

Huge result spaces could be navigated fast using our decision tree quality plot.

@@

Dented vertebrae indicate pathological deformation, and can be captured by segmenting the top and bottom point of each vertebra center.

@@

Combining the power of statistical analyses, visual analytics and classification techniques is essential for analyzing increasingly complex heterogenous population data.

@@

Our method provides a novel way to gain insight into these complex data sets and amplifies \emph{hypotheses generation}.

@@

\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics. The work was partly funded by the German Ministry of Education and Research (BMBF) within the STIMULATE project (grant no. 03FO16101A).
